### 2023-09-08 paper/auto_generated_draft.tex
auto generated paper drafts
These drafts were auto-generated by GPT-4-0314 based on source material
and will serve as a skeleton for me to edit later on.

\documentclass{article}
\usepackage[utf8]{inputenc}
\usepackage{amsmath, amsthm, amssymb}
\usepackage{graphicx}
\usepackage{algorithm}
\usepackage{algpseudocode}

\title{A Lock-Free Interaction Combinator Evaluator}
\author{Victor Taelin}
\date{}

\begin{document}

\maketitle

\begin{abstract}
We present a lock-free evaluator for interaction combinators, a simple yet powerful model of distributed computation. The proposed algorithm is based on the concept of implicit ownership regions and redirection wires, allowing for efficient parallel evaluation without race conditions or deadlocks. We provide detailed pseudocode for the core procedures and discuss their properties. Our implementation demonstrates promising performance and near-ideal speedup on massively parallel architectures.
\end{abstract}

\section{Introduction}\label{s:introduction}

Interaction nets are a graphical model of computation designed as a generalization of the proof structures of linear logic~\cite{girard}. They consist of graph-like structures composed of agents and edges, where each agent has a type and an arity connected to other agents via its ports. Interaction nets are inherently distributed, allowing computations to take place simultaneously in many parts of a net without synchronization. This makes them suitable for modeling massive parallelism.

A specific type of interaction nets is the system called \textit{interaction combinators}, introduced by Lafont~\cite{lafont}. These are built from three types of nodes: constructor (CON), duplicator (DUP), and eraser (ERA). Each node has a principal port and two auxiliary ports. The nodes interact according to six rules, divided into two categories: commutation and annihilation. Commutation rules apply when two different types of nodes are connected through their principal ports, while annihilation rules apply when two nodes of the same type are connected through their principal ports.

Interaction combinators possess several key properties such as determinism, locality, strong confluence (diamond property), and universality. These properties make them particularly suitable for modeling distributed and parallel computations~\cite{lafont}.

Despite the theoretical appeal of interaction combinators as a model of distributed computation, there is still no real-world implementation that performs well in practice 26 years after their introduction. In this paper, we propose a lock-free evaluator for interaction combinators that achieves massive parallelism on GPUs with promising performance and near-ideal speedup. The heart of our implementation is a lock-free \texttt{link()} procedure which we describe in detail.

The rest of the paper is organized as follows: Section~\ref{s:background} provides background information on interaction nets and interaction combinators; Section~\ref{s:lockfree} presents our lock-free evaluator proposal along with its core procedures; Section~\ref{s:pseudocode} provides detailed pseudocode for all functions; Finally, Section~\ref{s:conclusion} concludes the paper.

\section{Background}\label{s:background}

In this section, we provide an overview of interaction nets and interaction combinators to establish the necessary background for understanding our proposed algorithm.

\subsection{Interaction Nets}\label{ss:intnets}

Interaction nets were introduced by Lafont as a generalization of Girard's proof structures for linear logic \cite{girard}. They are graph-like structures composed of agents (nodes) connected by edges (wires). Each agent has a type and an arity determined by its number of ports. Ports can be connected to other agents via edges or left unconnected (free ports). The free ports form the interface of an interaction net.

The computation in an interaction net is performed by applying a set of \textit{interaction rules} that rewrite pairs of connected agents. These rules are local, meaning they only affect a small, constant-size region of the net. This locality property ensures that computations can take place simultaneously in many parts of an interaction net without synchronization, making them suitable for modeling parallelism.

\subsection{Interaction Combinators}\label{ss:intcombs}

Interaction combinators are a specific type of interaction nets introduced by Lafont \cite{lafont}. They consist of three types of nodes: constructor (CON), duplicator (DUP), and eraser (ERA). Each node has a principal port and two auxiliary ports. The nodes interact according to six rules, divided into two categories: commutation and annihilation. Commutation rules apply when two different types of nodes are connected through their principal ports, and annihilation rules apply when two nodes of the same type are connected through their principal ports.

\section{Lock-Free Evaluator}\label{s:lockfree}

In this section, we present our lock-free evaluator for interaction combinators. We first recap the basics of interaction combinators and discuss the problem with naive parallel reduction. Then, we briefly describe a lock-based solution and its limitations before presenting our lock-free proposal.

\subsection{Problem with Naive Parallel Reduction}\label{ss:problem}

The naive approach to reducing interaction combinators in parallel is to maintain a set of redexes (active pairs) and process them concurrently. However, this can lead to race conditions due to overlapping regions.

Consider the following example:

\begin{verbatim}
(CON x a a)
(CON x b c)
(CON y b c)
(CON y d d)
\end{verbatim}

There are two redexes: one formed by the first two nodes and another formed by the last two nodes. Each redex can affect two ports on the other redex (the 'b' and 'c' ports), resulting in an overlap of 2 ports among 12 total ports. Attempting to reduce these nodes in parallel naively could result in race conditions.

\subsection{Lock-Based Solution}\label{ss:lockbased}

A possible solution is to use locks on regions that need to be rewritten. For example, we could lock "surrounding" nodes before performing an interaction since a full rewrite can involve up to 6 nodes. Locking these nodes would make it thread-safe.

However, safely locking these 6 nodes presents challenges such as avoiding deadlocks or ensuring correct surrounding node sets after central node modifications by other threads.

One approach is first locking central nodes, which fixes surrounding node sets, then attempting surrounding node locks afterward. If successful, perform rewrite; otherwise release all locks and back-off for later retry attempts.

This solution works but may result in temporary deadlocks where threads repeatedly lock/unlock their respective regions until one obtains entire region control—performance heavily relies on the scheduler. On CPU, this algorithm performs satisfactorily; on GPU, where warp threads are in lockstep, performance may suffer.

\subsection{Lock-Free Proposal}\label{ss:proposal}

We propose an efficient lock-free reduction algorithm based on implicit ownership regions and redirection wires. The idea is to process a set of redexes in parallel but perform rewrites locally without immediately affecting surrounding nodes. Redirection wires allow us to replace central node ports with redirection pointers, enabling nodes pointing to them now to semantically point to their post-reduction locations.

When a redirection pointer points to a main port, we also implicitly own that pointer and start the \texttt{link()} procedure (detailed pseudocode provided in Section~\ref{s:pseudocode}). This procedure attempts connecting owned node (\texttt{src\_ref}) towards target location (\texttt{dir\_ptr}), resolving all encountered redirection nodes using atomic operations if necessary.

As long as the \texttt{link()} operation remains atomic, other parts of the algorithm can be performed in parallel. Moreover, we can break redexes into four segments for concurrent processing by separate threads:

\begin{itemize}
    \item Thread A1: moves A.aux1 towards B.aux1
    \item Thread A2: moves A.aux2 towards B.aux2
    \item Thread B1: moves B.aux1 towards A.aux1
    \item Thread B2: moves B.aux2 towards A.aux2
\end{itemize}

This optimization increases granularity further and allows 4x more cores than active redex count usage.

\section{Pseudocode}\label{s:pseudocode}

In this section, we provide detailed pseudocode for our proposed lock-free evaluator's core functions:

\begin{algorithm}
\caption{The link() procedure}\label{alg:link}
\begin{algorithmic}[1]
\Procedure{link}{$src\_ref: \&\text{Pointer}, dir\_ptr: \text{Pointer}$}
  \While{True}
    \State $trg\_ref \gets dir\_ptr.\text{target}()$
    \State $trg\_ptr \gets trg\_ref.\text{read}()$
    \If{$trg\_ptr.\text{is\_red}()$}
      \State $trg\_ref.\text{write}(0)$
      \State $dir\_ptr \gets trg\_ptr$
      \State \textbf{continue}
    \ElsIf{$trg\_ptr.\text{is\_aux}()$}
      \State $src\_ptr \gets src\_ref.\text{read}()$
      \If{$trg\_ref.\text{compare\_exchange}(trg\_ptr, src\_ptr) = trg\_ptr$}
        \State $trg\_ref \gets dir\_ptr.\text{target}()$
        \State $trg\_ptr \gets trg\_ref.\text{read}()$
        \While{$trg\_ptr.\text{is\_red}()$}
          \State $trg\_ref.\text{write}(0)$
          \State $trg\_ref \gets trg\_ptr.\text{target}()$
          \State $trg\_ptr \gets trg\_ref.\text{read}()$
        \EndWhile
        \State $src\_ref.\text{write}(0)$
        \State \textbf{return}
      \Else
        \State \textbf{continue}
      \EndIf
    \ElsIf{$trg\_ptr.\text{is\_nod}()$ or $trg\_ptr.\text{is\_tmp}()$}
      \State $fst\_ref, snd\_ref \gets \text{sort}(src\_ref, trg\_ref)$
      \State $fst\_ptr \gets fst\_ref.\text{exchange}(\text{TMP})$
      \If{not $fst\_ptr.\text{is\_tmp}()$}
        \State $snd\_ptr \gets snd\_ref.\text{exchange}(\text{TMP})$
        \State $\text{add\_redex}(fst\_ptr, snd\_ptr)$
        \State \textbf{return}
      \Else
        \State $fst\_ref.\text{write}(0)$
        \While{$snd\_ref.\text{compare\_exchange}(snd\_ptr, 0) \neq snd\_ptr$}
          \State \textbf{continue}
        \EndWhile
        \State \textbf{return}
      \EndIf
    \Else
      \State \textbf{continue}
    \EndIf
  \EndWhile
\EndProcedure
\end{algorithmic}
\end{algorithm}

\begin{algorithm}
\caption{The rewrite() function}\label{alg:rewrite}
\begin{algorithmic}[1]
\Procedure{rewrite}{$net$}
    \State \textbf{parallel for} $(a, b) \in net.\text{redexes}$
        \State $quarters \gets [(a, b, 1), (a, b, 2), (b, a, 1), (b, a, 2)]$
        \State \textbf{parallel for} $(a, b, port) \in quarters$
            \State $\text{match}(a\_ptr.\text{tag}(), b\_ptr.\text{tag}()):$
                \State \hspace{1cm}$(\text{CON}, \text{CON}) \Rightarrow \text{con\_con}(a, b, port)$
                \State \hspace{1cm}$(\text{CON}, \text{DUP}) \Rightarrow \text{con\_dup}(a, b, port)$
                \State \hspace{1cm}$(\text{CON}, \text{ERA}) \Rightarrow \text{con\_era}(a, b, port)$
                \State \hspace{1cm}$(\text{DUP}, \text{CON}) \Rightarrow \text{dup\_con}(b, a, port)$
                \State \hspace{1cm}$(\text{DUP}, \text{DUP}) \Rightarrow \text{dup\_dup}(b, b, port)$
                \State \hspace{1cm}$(\text{DUP}, \text{ERA}) \Rightarrow \text{dup\_era}(a, b, port)$
                \State \hspace{1cm}$(\text{ERA}, \text{CON}) \Rightarrow \text{era\_con}(b, a, port)$
                \State \hspace{1cm}$(\text{ERA}, \text{DUP}) \Rightarrow \text{era\_dup}(b, b, port)$
                \State \hspace{1cm}$(\text{ERA}, \text{ERA}) \Rightarrow \text{era\_era}(a, b, port)$
\EndProcedure
\end{algorithmic}
\end{algorithm}

\begin{algorithm}
\caption{The con\_con() function}\label{alg:con_con}
\begin{algorithmic}[1]
\Procedure{con\_con}{$a\_ptr: \&\text{Pointer}, b\_ptr: \&\text{Pointer}, port: \text{Port}$}
    \State $a\_aux\_ref \gets \&a\_ptr.\text{port}[port]$
    \State $b\_aux\_ref \gets \&b\_ptr.\text{port}[port]$
    \State $a\_aux\_ptr \gets a\_aux\_ref.\text{take}()$
    \State $b\_aux\_ref.\text{replace}(\text{TAKEN}, a\_aux\_ptr.\text{as\_redirection}())$
    \If{$a\_aux\_ptr.\text{is\_main}()$}
        \State $\text{link}(b\_aux\_ref, a\_ptr.\text{redirect}(port))$
    \EndIf
\EndProcedure
\end{algorithmic}
\end{algorithm}

\begin{algorithm}
\caption{The con\_dup() function}\label{alg:con_dup}
\begin{algorithmic}[1]
\Procedure{con\_dup}{$a\_ptr: \&\text{Pointer}, b\_ptr: \&\text{Pointer}, port: \text{Port}$}
    \State $a\_aux\_ref \gets \&a\_ptr.\text{port}[port]$
    \State $b\_aux\_ref \gets \&b\_ptr.\text{port}[port]$
    \State $a\_aux\_ptr \gets a\_aux\_ref.\text{take}()$
    \State $clone\_loc \gets \text{allocate\_location}()$
    \State $clone\_ptr \gets \text{mkptr}(\text{tag}(a\_ptr), clone\_loc)$
    \State $clone\_ptr.\text{fill\_inner\_wires}()$
    \State $b\_aux\_ref.\text{replace}(\text{TAKEN}, clone\_ptr)$
    \If{$a\_aux\_ptr.\text{is\_main}()$}
        \State $\text{link}(a\_aux\_ref, a\_aux\_ptr.\text{as\_redirection}())$
    \EndIf
    \If{$a\_aux\_ptr.\text{is\_main}()$}
        \State $\text{create\_redex}(a\_aux\_ptr, a\_aux\_ref.\text{take}())$
        \State $a\_aux\_ref.\text{compare\_exchange}(\text{TAKEN}, 0)$
    \EndIf
\EndProcedure
\end{algorithmic}
\end{algorithm}

\section{Conclusion}\label{s:conclusion}

We have presented a lock-free evaluator for interaction combinators based on implicit ownership regions and redirection wires. Our proposed algorithm allows efficient parallel evaluation without race conditions or deadlocks. The detailed pseudocode provided demonstrates the core procedures of our implementation.

Our lock-free evaluator shows promising performance and near-ideal speedup on massively parallel architectures such as GPUs. This work contributes to the practical realization of interaction combinators as a powerful model of distributed computation.

\bibliographystyle{plain}
\bibliography{references}

\end{document}

### 2023-09-08 cuda/hvm2.cu
fix aloc typo

@@ -258,7 +258,7 @@ __device__ inline Ptr* at(Net* net, Val idx, Port port) {
 __device__ inline u64 alloc(Worker *worker, Net *net) {
   u64 K = 0;
   while (true) {
-    u64  idx = (worker->aloc * 4 + worker->type) % NODE_SIZE;
+    u64  idx = worker->aloc % NODE_SIZE;
     u64* ref = (u64*)&net->node[idx].ports[P1];
     u64  got = atomicCAS(ref, 0, NEO);
     if (got == 0) {

### 2023-09-08 rust/src/core.rs

@@ -12,7 +12,7 @@
 // This expansion is performed on demand, and ERA-REF pointers are collected, allowing the runtime
 // to compute tail-recursive functions with constant memory usage.
 
-use std::{collections::HashMap, hash::BuildHasherDefault};
+use std::collections::HashMap;
 
 pub type Tag = u16;
 pub type Val = u64;
@@ -213,7 +213,7 @@ impl Net {
 
   // Allocates a consecutive chunk of 'size' nodes. Returns the index.
   #[inline(always)]
-  pub fn alloc(&mut self, size: usize) -> Val {
+  pub fn alloc(&mut self) -> Val {
     loop {
       if self.next >= self.node.len() {
         self.next = 0;
@@ -300,10 +300,10 @@ impl Net {
       self.free(b.val());
     // CON-DUP
     } else if a.is_ctr() && b.is_ctr() && a.tag() != b.tag() {
-      let x1 = self.alloc(1);
-      let x2 = self.alloc(1);
-      let y1 = self.alloc(1);
-      let y2 = self.alloc(1);
+      let x1 = self.alloc();
+      let x2 = self.alloc();
+      let y1 = self.alloc();
+      let y2 = self.alloc();
       self.set(x1, P1, Ptr::new(VR1, y1));
       self.set(x1, P2, Ptr::new(VR1, y2));
       self.set(x2, P1, Ptr::new(VR2, y1));
@@ -350,7 +350,7 @@ impl Net {
         // Allocates enough space...
         for i in 0 .. got.node.len() {
           unsafe {
-            *self.locs.get_unchecked_mut(i) = self.alloc(1);
+            *self.locs.get_unchecked_mut(i) = self.alloc();
           }
         }
         // Loads nodes, adjusting locations...

### 2023-09-08 rust/src/lang.rs

@@ -324,7 +324,7 @@ pub fn alloc_ltree(net: &mut Net, tree: &LTree, vars: &mut HashMap<String, Paren
       Ptr::new(ERA, 0)
     },
     LTree::Nod { tag, lft, rgt } => {
-      let val = net.alloc(1);
+      let val = net.alloc();
       let p1 = alloc_ltree(net, &*lft, vars, Parent::Node { val, port: P1 });
       net.set(val, P1, p1);
       let p2 = alloc_ltree(net, &*rgt, vars, Parent::Node { val, port: P2 });

### 2023-09-08 rust/src/lib.rs

@@ -1,0 +1,5 @@
+pub mod core;
+pub mod lang;
+
+pub use crate::core::*;
+pub use crate::lang::*;

### 2023-09-08 rust/src/lang.rs
Expose LNet internal fields to build it programatically

@@ -65,8 +65,8 @@ type LActs = Vec<(LTree,LTree)>;
 
 #[derive(Debug)]
 pub struct LNet {
-  root: LTree,
-  acts: LActs,
+  pub root: LTree,
+  pub acts: LActs,
 }
 
 // Parser

### 2023-09-08 paper/draft.tex
lock-free IC evaluator - manual draft

\documentclass{article}
\usepackage[utf8]{inputenc}
\usepackage{amsmath, amsthm, amssymb}
\usepackage{graphicx}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{listings}
\usepackage{xcolor}
\usepackage[export]{adjustbox}

\definecolor{codegreen}{rgb}{0,0.6,0}
\definecolor{codegray}{rgb}{0.5,0.5,0.5}
\definecolor{codepurple}{rgb}{0.58,0,0.82}
\definecolor{backcolour}{rgb}{0.95,0.95,0.92}

\lstset{
    language=Python, 
    basicstyle=\ttfamily\small,
    commentstyle=\color{codegreen},
    keywordstyle=\color{magenta},
    numberstyle=\tiny\color{codegray},
    stringstyle=\color{codepurple},
    breakatwhitespace=false,         
    breaklines=true,                 
    captionpos=b,                    
    keepspaces=true,                 
    numbers=left,                    
    numbersep=5pt,                  
    showspaces=false,                
    showstringspaces=false,
    showtabs=false,                  
    tabsize=2,
    frame=single,
    backgroundcolor=\color{backcolour}
}

\title{A Lock-Free Interaction Combinator Evaluator}
\author{Victor Taelin}
\date{}

\begin{document}

\maketitle

\begin{abstract}
We present a lock-free evaluator for interaction combinators, a simple yet powerful model of distributed computation. The proposed algorithm is based on the concept of implicit ownership regions and redirectors, allowing for efficient parallel evaluation without race conditions or deadlocks. We provide detailed pseudocode for the core procedures and illustrations. We have implemented our lock-free evaluator in CUDA and achieved successful reductions of large graphs on GPUs with thousands of concurrent threads, demonstrating the practicality and efficiency of our approach.
\end{abstract}

\section{Introduction}\label{s:introduction}

Interaction nets are a graphical model of computation designed as a generalization of the proof structures of linear logic~\cite{girard1987linear}. They consist of graph-like structures composed of agents and edges, where each agent has a type and an arity connected to other agents via its ports. Interaction nets are inherently distributed, allowing computations to take place simultaneously in many parts of a net without synchronization. This makes them suitable for modeling massive parallelism.

A specific type of interaction nets is the system called \textit{interaction combinators}, introduced by Lafont~\cite{lafont1997interaction}. These are built from three types of nodes: constructor (CON), duplicator (DUP), and eraser (ERA). Each node has a principal port and two auxiliary ports. The nodes interact according to six rules, divided into two categories: commutation and annihilation. Commutation rules apply when two different types of nodes are connected through their principal ports, while annihilation rules apply when two nodes of the same type are connected through their principal ports. In this paper, we use the symmetric interaction combinator variant, which has been described by Mazza~\cite{mazza2007denotational}, and is illustrated below:

\includegraphics[width=1.0\textwidth,margin=4 4 4 4]{interaction_rules.png}

Interaction combinators possess several key properties such as determinism, locality, strong confluence (diamond property), and universality. These properties make them particularly suitable for modeling distributed and parallel computations~\cite{lafont1997interaction}. Yet, despite that theoretical appeal, there has been limited practical exploration of their potential for massive parallel computation on modern hardware. In this paper, we propose a lock-free evaluator for interaction combinators that achieves near-ideal speedup on modern GPUs, with several thousand concurrent threads. The heart of our implementation is a lock-free \texttt{link()} procedure, which we describe in detail.

The paper is organized as follows: Section 2 discusses the limitations of a lock-based evaluator. Section 3 introduces our proposed lock-free evaluator. Section 4 presents an optimization to further increase parallelism. Section 5 discusses the implementation details and the performance of our CUDA-based evaluator~\cite{taelin2022lock}. Finally, Section 6 concludes the paper and discusses future work.

\section{Former Work}\label{s:former_work}

A lock-based evaluator for interaction nets was developed by Sato et al. in their work on Inpla: Interaction Nets as a Programming Language~\cite{hassan2010implementation, sato2014implementation}. Inpla is a multi-threaded parallel interpreter for interaction nets that aims to provide efficient execution of programs in both sequential and parallel environments. Inpla, based on the Lightweight calculus, focuses on the representation, calculus, data structures, and low-level language for implementing interaction nets. Their approach involves using compare-and-swap atomic operations and locking mechanisms, which can be sub-optimal when dealing with parallelism. The locking mechanisms used in Inpla can potentially lead to performance bottlenecks, especially when the checking process for locks spreads globally across the net.

Another approach, the ingpu evaluator~\cite{jiresch2014towards} was introduced as a GPU-based implementation of interaction nets. The ingpu evaluator, implemented using the CUDA/Thrust library, focused on the interaction and communication phases of the lightweight interaction calculus rather than maintaining an array of redexes and performing explicit graph rewrites. However, the ingpu method faced performance bottlenecks due to sorting and merging arrays, difficulty in representing the irregular graph structure, and varying output size of a reduction.

\section{Lock-Based Evaluator}\label{s:lockfree}

The naive approach to reducing interaction combinators in parallel is to represent nets as vectors of nodes, with ports holding pointers to their destinations, and maintain a set of redexes (active pairs) to process concurrently. However, this can lead to race conditions due to overlapping regions. To address this, the affected regions, which include two active nodes and up to 4 surrounding nodes, must be locked. This approach is illustrated below:

\includegraphics[width=1.0\textwidth,margin=4 4 4 4]{lock_based_evaluator_1.png}
\includegraphics[width=1.0\textwidth,margin=4 4 4 4]{lock_based_evaluator_2.png}
\includegraphics[width=1.0\textwidth,margin=4 4 4 4]{lock_based_evaluator_3.png}

While this strategy works, locking the 6 surrounding nodes can be challenging, potentially leading to a deadlock. For example, consider thread R locking active nodes A and B, while thread G locks active nodes C and D. To proceed with the reduction, thread R must lock C, and thread G must lock B, which is impossible in this situation.

To avoid deadlocks, one could attempt to lock nodes in an established order, such as from left to right. This prevents deadlocks but may cause a thread to lock an irrelevant region. For instance, suppose thread R locks X and Y immediately before another thread reduces the X node, replacing it with K. Thread R then locks A, B, and C, incorrectly locking the region X, Y, A, B, and C, instead of including the new node K. Consequently, thread R cannot proceed with its reduction.

Any viable solution to this issue will inevitably involve back-offs or temporary deadlocks, causing the algorithm to rely on the scheduler. This dependence can have negative implications for performance, particularly in architectures such as GPUs, where warps operate in lockstep. The tight synchronization in such systems can exacerbate performance issues, leading to less efficient execution of the algorithm.

\section{Lock-Free Evaluator}\label{s:lockfree}

In this section, we propose an efficient lock-free reduction algorithm based on implicit ownership regions and redirection wires. The idea is to process a set of redexes in parallel, but perform rewrites only within implicitly owned regions, eliminating the need for locks.

When a thread obtains a redex, it assumes ownership of its two active nodes. It then replaces these nodes with redirection wires, allowing the redex to be semantically reduced without affecting the surrounding region. Next, the thread expands its ownership region to include surrounding nodes connected to it via main ports. Finally, the thread invokes a linking procedure that starts from these surrounding main ports and traverses the graph, clearing redirection wires until arriving at a port outside its ownership region. At this point, there are two cases to consider:

1. If the outside port is an auxiliary port, the thread attempts to connect the surrounding main port to it using an atomic compare-exchange operation and clears the backward path.

2. If it is a main port, the thread has found a new active pair and coordinates with the opposing thread — which will traverse the backward path — to create a new redex.

Below is a complete illustration of this process:

\includegraphics[width=1.0\textwidth,margin=4 4 4 4]{lock_free_evaluator_1.png}
\includegraphics[width=1.0\textwidth,margin=4 4 4 4]{lock_free_evaluator_2.png}
\includegraphics[width=1.0\textwidth,margin=4 4 4 4]{lock_free_evaluator_3.png}
\includegraphics[width=1.0\textwidth,margin=4 4 4 4]{lock_free_evaluator_4.png}
\includegraphics[width=1.0\textwidth,margin=4 4 4 4]{lock_free_evaluator_5.png}
\includegraphics[width=1.0\textwidth,margin=4 4 4 4]{lock_free_evaluator_6.png}
\includegraphics[width=1.0\textwidth,margin=4 4 4 4]{lock_free_evaluator_7.png}

Note that in Image 2, the reduction is semantically complete, with ports X1-W0 and Y0-Z0 indirectly connected. Each thread independently performs this reduction, without impacting surrounding nodes, by converting the auxiliary ports of their owned nodes to redirectors and then swapping opposing ports (e.g., A1 goes to B1, B2 goes to A2, etc.). This approach entirely avoids locks. The link() procedure is then only needed to clear the memory occupied by redirectors and to identify new active pairs. As this clean-up process may cross ownership regions, caution is required to deploy atomics when necessary. The following is a comprehensive pseudocode:

\begin{lstlisting}
# Atomically links the node in 'src_ref' towards 'dir_ptr'.
def link(src_ref: &Pointer, dir_ptr: Pointer):
  while True:
    # Peek the target, which may not be owned by us.
    trg_ref = dir_ptr.target()
    trg_ptr = trg_ref.read()

    # If target is a redirection, clear and move forward.
    if trg_ptr.is_red():

      # We own the redirection, so we can mutate it.
      trg_ref.write(0)
      dir_ptr = trg_ptr
      continue

    # If target is an aux port, try replacing it with the node.
    elif trg_ptr.is_aux():

      # Peeks the source node.
      src_ptr = src_ref.read()

      # We don't own the port, so we try replacing it.
      if trg_ref.compare_exchange(trg_ptr, src_ptr) == trg_ptr:
        # Collect the orphaned backward path.
        trg_ref = dir_ptr.target()
        trg_ptr = trg_ref.read()
        while trg_ptr.is_red():
          trg_ref.write(0)
          trg_ref = trg_ptr.target()
          trg_ptr = trg_ref.read()

        # Clear source location.
        src_ref.write(0)
        return

      # If the compare_exchange failed, we try again.
      else:
        continue

    # If target is a main port, two threads reach this branch.
    elif trg_ptr.is_nod() or trg_ptr.is_tmp():

      # Sort references, to avoid deadlocks.
      fst_ref, snd_ref = sort(src_ref, trg_ref)

      # Swap the first reference by TMP placeholder.
      fst_ptr = fst_ref.exchange(TMP)

      # First to arrive creates a redex.
      if !fst_ptr.is_tmp():
        snd_ptr = snd_ref.exchange(TMP)
        add_redex(fst_ptr, snd_ptr)
        return

      # Second to arrive clears up the memory.
      else:
        fst_ref.write(0)
        while snd_ref.compare_exchange(snd_ptr, 0) != snd_ptr:
          continue
        return

    # If it is taken, we wait.
    else:
      continue
\end{lstlisting}

\section{Optimization: 1/4 rewrites}\label{s:lockfree}

The lock-free evaluator procedure above reveals a symmetry that can be exploited to further increase the parallelism of the algorithm. Notice how thread R performs the same logic 4 times:

- Move A1 as redirector to B1; if target is main, link B1 towards A1.

- Move A2 as redirector to B2; if target is main, link B2 towards A2.

- Move B1 as redirector to A1; if target is main, link A1 towards B1.

- Move B2 as redirector to A2; if target is main, link A2 towards B2.

These 4 segments are, themselves, independent, and can be performed in parallel. This allows a single interaction to be processed by 4 threads, each one performing 1/4 of a rewrite. According to Amdahl's law, the maximum speedup achievable through parallelization is limited by the fraction of the program that must be executed sequentially. By dividing the rewrite process into 4 smaller sequential chunks, we effectively reduce the sequential portion of the algorithm, increasing the maximum potential speedup by 4x. The code below demonstrates the implementation of this optimized approach, completing the algorithm and allowing for efficient parallel execution of interaction combinators.

\begin{lstlisting}
# Performs 1/4 con-con interaction
def con_con(a_ptr: &Pointer, b_ptr: &Pointer, port: Port):

  # Gets a reference current and opposing ports
  a_aux_ref = &a_ptr.port[port]
  b_aux_ref = &b_ptr.port[port]

  # Takes the current port and casts to redirector
  a_aux_ptr = a_aux_ref.exchange(0).as_redirector()

  # Synchronizes local threads
  local_threads.sync()

  # Sends current port to opposing port
  b_aux_ref.write(a_aux_ptr)

  # Synchronizes local threads
  local_threads.sync()

  # If the current port targeted a main port...
  if a_aux_ptr.targets_main():

    # Link the opposing port towards the current port
    link(b_aux_ref, new_ptr_to(a_aux_ptr))

# Performs 1/4 con-dup interaction
def con_dup(a_ptr: &Pointer, b_ptr: &Pointer, port: Port):

  # Gets reference to both aux ptrs
  a_aux_ref = &a_ptr.port[port]
  b_aux_ref = &b_ptr.port[port]

  # Takes my aux ptr
  a_aux_ptr = a_aux_ref.exchange(0).as_redirector()

  # Allocates a new clone
  clone_loc = malloc(1 * size(Ptr))
  clone_ptr = mkptr(a_ptr.tag(), clone_loc)

  # Synchronizes local threads
  local_threads.sync()

  # Communicates this clone's loc to local threads
  local_threads.send_clone_loc(clone_loc)

  # Synchronizes local threads
  local_threads.sync()

  # Gets opposing clone locs
  clone_x_loc, clone_y_loc = local_threads.receive_clone_locs()

  # Fills clone inner wires
  clone_loc[1] = clone_x_loc
  clone_loc[2] = clone_y_loc

  # Sends clone to opposing port
  b_aux_ref.write(clone_ptr)

  # Synchronizes local threads
  local_threads.sync()

  # If the current port targeted an aux port...
  if a_aux_ptr.targets_aux():

    # Link the current port towards its former target
    link(a_aux_ref, a_aux_ptr)

  # If the current port targeted a main port...
  if a_aux_ptr.is_main():

    # Form a new redex between its former and current target
    create_redex(a_aux_ptr, a_aux_ref.exchange(0))
\end{lstlisting}

\begin{lstlisting}
def rewrite(net):
  # Performs each interaction, in parallel
  p-for (a,b) in net.redexes:

    # Sets up the 1/4 interactions
    quarters = [(a,b,1), (a,b,2), (b,a,1), (b,a,2)]

    # Performs each 1/4 interaction, in parallel
    p-for (a, b, port) in quarters:
      match (a_ptr.tag(), b_ptr.tag()):
        (CON, CON) => con_con(a, b, port)
        (CON, DUP) => con_dup(a, b, port)
        (CON, ERA) => con_era(a, b, port)
        (DUP, CON) => dup_con(b, a, port)
        (DUP, DUP) => dup_dup(b, b, port)
        (DUP, ERA) => dup_era(a, b, port)
        (ERA, CON) => era_con(b, a, port)
        (ERA, DUP) => era_dup(b, b, port)
        (ERA, ERA) => era_era(a, b, port)
\end{lstlisting}

\section{Implementation}\label{s:implementation}

The lock-free evaluator algorithm has been implemented in CUDA~\cite{taelin2022lock} and has successfully reduced large graphs on GPUs with thousands of concurrent threads, suggesting the algorithm's correctness, even though a formal proof is not provided yet. Moreover, our CUDA implementation demonstrates efficient performance, reaching 900 million rewrites per second on an RTX 4090, which is 15 times more than a single-core CPU evaluator. This significant speedup confirms the practical applicability and efficiency of our approach.

\section{Conclusion}\label{s:conclusion}

In this paper, we have presented a lock-free evaluator for interaction combinators that achieves near-ideal speedup on modern GPUs. Our approach leverages the inherent parallelism of interaction combinators and avoids the pitfalls of lock-based evaluators. We also introduced an optimization that further increases the potential for parallelism by breaking down rewrites into smaller sequential chunks. Our CUDA implementation demonstrates the practicality and efficiency of our approach, as well as suggesting its correctness. Future work includes further optimizations, exploring other applications of our lock-free evaluator, and providing formal proofs of correctness.

\bibliographystyle{plain}
\bibliography{references}

\begin{thebibliography}{9}
\bibitem{girard1987linear} 
Girard, J. Y. (1987). Linear logic. Theoretical computer science, 50(1), 1-101.

\bibitem{lafont1997interaction} 
Lafont, Y. (1997). Interaction combinators. Information and Computation, 137(1), 69-101.

\bibitem{jiresch2014towards} 
Jiresch, E. (2014). Towards a GPU-based implementation of interaction nets. arXiv preprint arXiv:1404.0076.

\bibitem{mazza2007denotational} 
Mazza, D. (2007). A denotational semantics for the symmetric interaction combinators. Mathematical Structures in Computer Science, 17(3), 527-562.

\bibitem{sato2014implementation}
Sato, S. (2014). Design and implementation of a low-level language for interaction nets. Ph.D. Thesis, University of Sussex.

\bibitem{taelin2022lock} 
Taelin, V. (2022). Higher-order Virtual Machine [Online]. Available at: https://github.com/HigherOrderCO/hvm-core

\end{document}

### 2023-09-08 cuda/hvm2.cu

@@ -108,15 +108,15 @@ typedef struct {
 
 // A worker local data
 typedef struct {
-  u64   tid;   // thread id
-  u64   bid;   // block id 
-  u64   gid;   // global id
-  u64   type;  // worker type (A1|A2|B1|B2)
-  u64   port;  // worker port (P1|P2)
-  u64   aloc;  // where to alloc next node
-  u64   rwts;  // local rewrites performed
-  Wire* bag;   // local redex bag
-  u64*  locs;  // local alloc locs
+  u64   tid;  // thread id
+  u64   bid;  // block id 
+  u64   gid;  // global id
+  u64   type; // worker type (A1|A2|B1|B2)
+  u64   port; // worker port (P1|P2)
+  u64   aloc; // where to alloc next node
+  u64   rwts; // local rewrites performed
+  Wire* bag;  // local redex bag
+  u64*  locs; // local alloc locs
 } Worker;
 
 // Debug
@@ -3318,3 +3318,37 @@ int main() {
 
   return 0;
 }
+
+// Possible issues (auto-detected):
+// 1. In the function `rng`, there is no seed provided for the random number generator. This could lead to generating the same sequence of numbers every time the program runs.
+// 2. In the function `mkptr`, there is no check if the value `val` exceeds the limit of 48 bits. If it does, this would lead to incorrect results due to overflow.
+// 3. In the function `div`, there is no check for division by zero. This could lead to a runtime error.
+// 4. In the function `alloc`, there is an infinite loop with no exit condition if all nodes are occupied. This could lead to a deadlock.
+// 5. In the function `replace`, there is an infinite loop if the expected value never matches the current value at the reference. This could lead to a deadlock.
+// 6. In the function `link`, there is no check if the target reference is NULL before dereferencing it. This could lead to a segmentation fault.
+// 7. In the function `scansum`, there is no check if the array size is a power of two. If it's not, this could lead to incorrect results or out-of-bounds access.
+// 8. In the function `global_scatter_prepare_0`, there is no check if the block index exceeds the maximum block size. This could lead to out-of-bounds access.
+// 9. In the function `global_rewrite`, there is no check if the thread id exceeds the maximum thread count. This could lead to out-of-bounds access.
+// 10. In the function `do_global_scatter`, there is no check if the total length of bags exceeds the maximum bag size. This could lead to out-of-bounds access.
[…]

### 2023-09-08 paper/draft.tex
fix bibliography
paper formatting improvements

@@
     frame=single,
-    backgroundcolor=\color{backcolour}
+    backgroundcolor=\color{white}
 }
@@
 The proposed algorithm is based on the concept of implicit ownership regions and redirectors, allowing for efficient parallel evaluation without race conditions or deadlocks.
-We provide detailed pseudocode for the core procedures and illustrations.
+We provide detailed illustrations, and the pseudocode for the core procedures can be found in the appendix.
 We have implemented our lock-free evaluator in CUDA and achieved successful reductions of large graphs on GPUs with thousands of concurrent threads, demonstrating the practicality and efficiency of our approach.
@@
 
+\begin{figure}[h!]
+\centering
 \includegraphics[width=1.0\textwidth,margin=4 4 4 4]{interaction_rules.png}
+\caption{Interaction rules}
+\end{figure}
 
@@
 
+\begin{figure}[h!]
+\centering
 \includegraphics[width=1.0\textwidth,margin=4 4 4 4]{lock_based_evaluator_1.png}
+\caption{Lock-based evaluator step 1}
+\end{figure}
+
+\begin{figure}[h!]
+\centering
 \includegraphics[width=1.0\textwidth,margin=4 4 4 4]{lock_based_evaluator_2.png}
+\caption{Lock-based evaluator step 2}
+\end{figure}
+
+\begin{figure}[h!]
+\centering
 \includegraphics[width=1.0\textwidth,margin=4 4 4 4]{lock_based_evaluator_3.png}
+\caption{Lock-based evaluator step 3}
+\end{figure}
 
@@
 
-Below is a complete illustration of this process:
+The following illustrations showcase this process:
 
+\begin{figure}[h!]
+\centering
 \includegraphics[width=1.0\textwidth,margin=4 4 4 4]{lock_free_evaluator_1.png}
+\caption{Lock-free evaluator step 1}
+\end{figure}
+
+\begin{figure}[h!]
+\centering
 \includegraphics[width=1.0\textwidth,margin=4 4 4 4]{lock_free_evaluator_2.png}
+\caption{Lock-free evaluator step 2}
+\end{figure}
+
+\begin{figure}[h!]
+\centering
 \includegraphics[width=1.0\textwidth,margin=4 4 4 4]{lock_free_evaluator_3.png}
+\caption{Lock-free evaluator step 3}
+\end{figure}
+
+\begin{figure}[h!]
+\centering
 \includegraphics[width=1.0\textwidth,margin=4 4 4 4]{lock_free_evaluator_4.png}
+\caption{Lock-free evaluator step 4}
+\end{figure}
+
+\begin{figure}[h!]
+\centering
 \includegraphics[width=1.0\textwidth,margin=4 4 4 4]{lock_free_evaluator_5.png}
+\caption{Lock-free evaluator step 5}
+\end{figure}
+
+\begin{figure}[h!]
+\centering
 \includegraphics[width=1.0\textwidth,margin=4 4 4 4]{lock_free_evaluator_6.png}
+\caption{Lock-free evaluator step 6}
+\end{figure}
+
+\begin{figure}[h!]
+\centering
 \includegraphics[width=1.0\textwidth,margin=4 4 4 4]{lock_free_evaluator_7.png}
+\caption{Lock-free evaluator step 7}
+\end{figure}
 
@@
 This approach entirely avoids locks.
-The link() procedure is then only needed to clear the memory occupied by redirectors and to identify new active pairs.
+The link() procedure in the appendix is then only needed to clear the memory occupied by redirectors and to identify new active pairs.
 As this clean-up process may cross ownership regions, caution is required to deploy atomics when necessary.
-The following is a comprehensive pseudocode:
+
+\section{Optimization: 1/4 rewrites}\label{s:lockfree}
+
+The lock-free evaluator procedure above reveals a symmetry that can be exploited to further increase the parallelism of the algorithm. Notice how thread R performs the same logic 4 times:
+
+\begin{itemize}
+  \item Move A1 as redirector to B1; if target is main, link B1 towards A1.
+  \item Move A2 as redirector to B2; if target is main, link B2 towards A2.
+  \item Move B1 as redirector to A1; if target is main, link A1 towards B1.
+  \item Move B2 as redirector to A2; if target is main, link A2 towards B2.
+\end{itemize}
+
+These 4 segments are, themselves, independent, and can be performed in parallel.
+This allows a single interaction to be processed by 4 threads, each one performing 1/4 of a rewrite.
+According to Amdahl's law, the maximum speedup achievable through parallelization is limited by the fraction of the program that must be executed sequentially.
+By dividing the rewrite process into 4 smaller sequential chunks, we effectively reduce the sequential portion of the algorithm, increasing the maximum potential speedup by 4x.
+The pseudocode for this optimized approach can be found in the appendix, completing the algorithm and allowing for efficient parallel execution of interaction combinators.
+
+\section{Implementation}\label{s:implementation}
+
+The lock-free evaluator algorithm has been implemented in CUDA~\cite{taelin2022lock} and has successfully reduced large graphs on GPUs with thousands of concurrent threads, suggesting the algorithm's correctness, even though a formal proof is not provided yet.
+Moreover, our CUDA implementation demonstrates efficient performance, reaching 900 million rewrites per second on an RTX 4090, which is 15 times more than a single-core CPU evaluator.
+This significant speedup confirms the practical applicability and efficiency of our approach.
+
+\section{Conclusion}\label{s:conclusion}
+
+In this paper, we have presented a lock-free evaluator for interaction combinators that achieves near-ideal speedup on modern GPUs.
+Our approach leverages the inherent parallelism of interaction combinators and avoids the pitfalls of lock-based evaluators.
+We also introduced an optimization that further increases the potential for parallelism by breaking down rewrites into smaller sequential chunks.
+Our CUDA implementation demonstrates the practicality and efficiency of our approach, as well as suggesting its correctness.
+Future work includes further optimizations, exploring other applications of our lock-free evaluator, and providing formal proofs of correctness.
+
+\begin{thebibliography}{9}
+\bibitem{girard1987linear} 
+Girard, J. Y. (1987). Linear logic. Theoretical computer science, 50(1), 1-101.
+
+\bibitem{lafont1997interaction} 
+Lafont, Y. (1997). Interaction combinators. Information and Computation, 137(1), 69-101.
+
+\bibitem{jiresch2014towards} 
+Jiresch, E. (2014). Towards a GPU-based implementation of interaction nets. arXiv preprint arXiv:1404.0076.
+
+\bibitem{mazza2007denotational} 
+Mazza, D. (2007). A denotational semantics for the symmetric interaction combinators. Mathematical Structures in Computer Science, 17(3), 527-562.
+
+\bibitem{sato2014implementation}
+Sato, S. (2014). Design and implementation of a low-level language for interaction nets. Ph.D. Thesis, University of Sussex.
+
+\bibitem{taelin2022lock} 
+Taelin, V. (2022). Higher-order Virtual Machine [Online]. Available at: https://github.com/HigherOrderCO/hvm-core
+\end{thebibliography}
+
+\appendix
+\section{Pseudocode}\label{app:pseudocode}
+
+In this appendix, we provide the complete pseudocode for the lock-free evaluator algorithm and the 1/4 rewrites optimization.
 
@@
 \end{lstlisting}
-
-\section{Optimization: 1/4 rewrites}\label{s:lockfree}
-
-The lock-free evaluator procedure above reveals a symmetry that can be exploited to further increase the parallelism of the algorithm. Notice how thread R performs the same logic 4 times:
-
-- Move A1 as redirector to B1; if target is main, link B1 towards A1.
-
-- Move A2 as redirector to B2; if target is main, link B2 towards A2.
-
-- Move B1 as redirector to A1; if target is main, link A1 towards B1.
-
-- Move B2 as redirector to A2; if target is main, link A2 towards B2.
-
-These 4 segments are, themselves, independent, and can be performed in parallel.
-This allows a single interaction to be processed by 4 threads, each one performing 1/4 of a rewrite.
-According to Amdahl's law, the maximum speedup achievable through parallelization is limited by the fraction of the program that must be executed sequentially.
-By dividing the rewrite process into 4 smaller sequential chunks, we effectively reduce the sequential portion of the algorithm, increasing the maximum potential speedup by 4x.
-The code below demonstrates the implementation of this optimized approach, completing the algorithm and allowing for efficient parallel execution of interaction combinators.
 
@@
 
-\section{Implementation}\label{s:implementation}
-
-The lock-free evaluator algorithm has been implemented in CUDA~\cite{taelin2022lock} and has successfully reduced large graphs on GPUs with thousands of concurrent threads, suggesting the algorithm's correctness, even though a formal proof is not provided yet.
-Moreover, our CUDA implementation demonstrates efficient performance, reaching 900 million rewrites per second on an RTX 4090, which is 15 times more than a single-core CPU evaluator.
-This significant speedup confirms the practical applicability and efficiency of our approach.
-
-\section{Conclusion}\label{s:conclusion}
-
-In this paper, we have presented a lock-free evaluator for interaction combinators that achieves near-ideal speedup on modern GPUs.
-Our approach leverages the inherent parallelism of interaction combinators and avoids the pitfalls of lock-based evaluators.
-We also introduced an optimization that further increases the potential for parallelism by breaking down rewrites into smaller sequential chunks.
-Our CUDA implementation demonstrates the practicality and efficiency of our approach, as well as suggesting its correctness.
-Future work includes further optimizations, exploring other applications of our lock-free evaluator, and providing formal proofs of correctness.
-
-\bibliographystyle{plain}
-\bibliography{references}
-
-\begin{thebibliography}{9}
-\bibitem{girard1987linear} 
-Girard, J. Y. (1987). Linear logic. Theoretical computer science, 50(1), 1-101.
-
-\bibitem{lafont1997interaction} 
-Lafont, Y. (1997). Interaction combinators. Information and Computation, 137(1), 69-101.
-
-\bibitem{jiresch2014towards} 
-Jiresch, E. (2014). Towards a GPU-based implementation of interaction nets. arXiv preprint arXiv:1404.0076.
-
-\bibitem{mazza2007denotational} 
-Mazza, D. (2007). A denotational semantics for the symmetric interaction combinators. Mathematical Structures in Computer Science, 17(3), 527-562.
-
-\bibitem{sato2014implementation}
-Sato, S. (2014). Design and implementation of a low-level language for interaction nets. Ph.D. Thesis, University of Sussex.
-
-\bibitem{taelin2022lock} 
-Taelin, V. (2022). Higher-order Virtual Machine [Online]. Available at: https://github.com/HigherOrderCO/hvm-core
-
 \end{document}

### 2023-09-08 paper/draft.tex
:|

@@
 Commutation rules apply when two different types of nodes are connected through their principal ports, while annihilation rules apply when two nodes of the same type are connected through their principal ports.
-In this paper, we use the symmetric interaction combinator variant, which has been described by Mazza~\cite{mazza2007denotational}, and is illustrated below:
+In this paper, we use the symmetric interaction combinator variant, which has been described by Mazza~\cite{mazza2007denotational}, and is illustrated below.
 

### 2023-09-10 rust/src/core.rs
CUDA local reductions - WIP

@@ -15,23 +15,21 @@
 use std::collections::HashMap;
 
 pub type Tag = u16;
-pub type Val = u64;
+pub type Val = u32;
 
 // Core terms
-pub const NIL: Tag = 0x0000; // empty node
-pub const REF: Tag = 0x0001; // reference to a definition (closed net)
-pub const ERA: Tag = 0x0002; // unboxed eraser
-pub const VRR: Tag = 0x0003; // variable pointing to root
-pub const VR1: Tag = 0x0004; // variable pointing to aux1 port of node
-pub const VR2: Tag = 0x0005; // variable pointing to aux2 port of node
-pub const RDR: Tag = 0x0006; // redirection to root
-pub const RD1: Tag = 0x0007; // redirection to aux1 port of node
-pub const RD2: Tag = 0x0008; // redirection to aux2 port of node
-pub const CON: Tag = 0x1000; // points to main port of con node
-pub const DUP: Tag = 0x1001; // points to main port of dup node; higher labels also dups
-
-// Numeric terms
-pub const NUM: Tag = 0x0100; // unboxed u48
+pub const NIL: Tag = 0x0; // empty node
+pub const REF: Tag = 0x1; // reference to a definition (closed net)
+pub const ERA: Tag = 0x2; // unboxed eraser
+pub const VRR: Tag = 0x3; // variable pointing to root
+pub const VR1: Tag = 0x4; // variable pointing to aux1 port of node
+pub const VR2: Tag = 0x5; // variable pointing to aux2 port of node
+pub const RDR: Tag = 0x6; // redirection to root
+pub const RD1: Tag = 0x7; // redirection to aux1 port of node
+pub const RD2: Tag = 0x8; // redirection to aux2 port of node
+pub const NUM: Tag = 0x9; // redirection to aux2 port of node
+pub const CON: Tag = 0xA; // points to main port of con node
+pub const DUP: Tag = 0xB; // points to main port of dup node; higher labels also dups
 
 // A node port: 1 or 2. Main ports are omitted.
 pub type Port = usize;
@@ -63,32 +61,32 @@ pub struct Net {
   pub root: Ptr,
   pub acts: Vec<(Ptr, Ptr)>,
   pub node: Vec<Node>,
-  pub head: Vec<Ptr>,
+  pub isnf: Vec<bool>,
   pub used: usize,
   pub rwts: usize,
       next: usize,
[…]

### 2023-09-10 rust/src/lang.rs

@@ -88,14 +88,14 @@ fn consume(chars: &mut Peekable<Chars>, text: &str) {
   }
 }
 
-pub fn parse_num_lit(chars: &mut Peekable<Chars>) -> u64 {
-  let mut num : u64 = 0;
+pub fn parse_num_lit(chars: &mut Peekable<Chars>) -> u32 {
+  let mut num : u32 = 0;
   skip_spaces(chars);
   while let Some(c) = chars.peek() {
     if !c.is_digit(10) {
       break;
     }
-    num = num * 10 + c.to_digit(10).unwrap() as u64;
+    num = num * 10 + c.to_digit(10).unwrap() as u32;
     chars.next();
   }
   num
@@ -133,7 +133,7 @@ pub fn parse_ltree(chars: &mut Peekable<Chars>) -> LTree {
       chars.next();
       skip_spaces(chars);
       let name = parse_str_lit(chars);
-      LTree::Ref { nam: name_to_u64(&name) }
+      LTree::Ref { nam: name_to_u32(&name) }
     },
     Some(c) if c.is_digit(10) => {
       LTree::NUM { val: parse_num_lit(chars) }
@@ -187,7 +187,7 @@ pub fn show_ltree(tree: &LTree) -> String {
       nam.clone()
     },
     LTree::Ref { nam } => {
-      format!("@{}", u64_to_name(*nam))
+      format!("@{}", u32_to_name(*nam))
     },
     LTree::NUM { val } => {
       val.to_string()
@@ -281,7 +281,7 @@ pub fn letters_to_name(letters: Vec<u8>) -> String {
   return name;
 }
 
-pub fn u64_to_letters(num: u64) -> Vec<u8> {
+pub fn u32_to_letters(num: u32) -> Vec<u8> {
   let mut letters = Vec::new();
   let mut num = num;
   while num > 0 {
@@ -292,20 +292,20 @@ pub fn u64_to_letters(num: u64) -> Vec<u8> {
   return letters;
 }
 
-pub fn letters_to_u64(letters: Vec<u8>) -> u64 {
+pub fn letters_to_u32(letters: Vec<u8>) -> u32 {
   let mut num = 0;
   for letter in letters {
-    num = num * 64 + letter as u64;
+    num = num * 64 + letter as u32;
   }
   return num;
 }
 
-pub fn name_to_u64(name: &str) -> u64 {
-  letters_to_u64(name_to_letters(name))
[…]

### 2023-09-10 rust/src/main.rs

@@ -184,14 +184,14 @@ fn main() {
   // ex0 = ((n S) Z)
   define(book, "ex0", "
     $ root
-    & @c2 ~ (0 @S (0 @Z root))
+    & @c2 ~ (0 @k2 root)
   ");
 
   // Allocates a big tree
   // ex1 = ((n g_s) g_z)
   define(book, "ex1", "
     $ root
-    & @c2 ~ (0 @g_s (0 @g_z root))
+    & @c23 ~ (0 @g_s (0 @g_z root))
   ");
 
   // This example decreases a binary counter until it reaches 0. It uses recursion, based on
@@ -204,47 +204,44 @@ fn main() {
   define(book, "ex2", "
     $ main
     & @run ~ (0 nie main)
-    & @c6  ~ (0 @I (0 @E nie))
+    & @c7  ~ (0 @I (0 @E nie))
   "); 
 
   // Decreases many binary counters until they reach 0
   define(book, "ex3", "
     $ res
     & @brn ~ (0 dep res)
-    & @c14 ~ (0 @S (0 @Z dep))
+    & @c4  ~ (0 @S (0 @Z dep))
   "); 
 
   // Initializes the net
   let net = &mut Net::new(1 << 26);
-  net.boot(name_to_u64("ex3"));
+  net.boot(book, name_to_u32("ex2"));
 
   // Computes its normal form
   net.normal(book);
 
   // Shows results and stats
-  //println!("[net]\n{}", show_net(&net));
+  println!("[net]\n{}", show_net(&net));
   //println!("{}", net.head.iter().map(|x| format!("{:016x}", x.data)).collect::<Vec<String>>().join(" "));
   println!("size: {}", net.node.len());
   println!("used: {}", net.used);
   println!("rwts: {}", net.rwts);
-
-  println!("{}", u64_to_name(0x36E72));
-  //Prints book to use on CUDA
-  //print_book_to_cuda(&book);
+  //populate_cuda(&book); // prints CUDA book
 }
 
-fn print_book_to_cuda(book: &Book) {
+fn populate_cuda(book: &Book) {
   println!("Term* term;");
-  let mut sorted_defs: Vec<(&u64, &Net)> = book.defs.iter().collect();
+  let mut sorted_defs: Vec<(&u32, &Net)> = book.defs.iter().collect();
   sorted_defs.sort_by_key(|&(key, _)| key);
   for (key, def) in sorted_defs {
-    println!("  // {}", u64_to_name(*key));
[…]

### 2023-09-11 cuda/hvm2.cu
CUDA local reductions - WIP

@@ -115,13 +115,13 @@ typedef struct {
   u32   aloc; // where to alloc next node
   u32   rwts; // local rewrites performed
   u32*  locs; // local alloc locs
-  u64*  self_rlen; // local redex bag length
-  Wire* self_rbag; // local redex bag
-  Wire* self_rpop; // popped redex
-  u64*  near_rlen; // local neighbor redex bag len
-  Wire* near_rbag; // local neighbor redex bag
-  u64*  afar_rlen; // global neighbor redex bag len
-  Wire* afar_rbag; // global neighbor redex bag
+  u64*  rlen; // local redex bag length
+  Wire* rbag; // local redex bag
+  Wire* rpop; // popped redex
+  //u64*  near_rlen; // local neighbor redex bag len
+  //Wire* near_rbag; // local neighbor redex bag
+  //u64*  afar_rlen; // global neighbor redex bag len
+  //Wire* afar_rbag; // global neighbor redex bag
 
 } Worker;
 
@@ -248,12 +248,12 @@ __host__ __device__ inline Wire mkwire(Ptr p1, Ptr p2) {
 }
 
 // Gets the left element of a wire
-__host__ __device__ inline Wire wire_lft(Wire wire) {
+__host__ __device__ inline Ptr wire_lft(Wire wire) {
   return wire >> 32;
 }
 
 // Gets the right element of a wire
-__host__ __device__ inline Wire wire_rgt(Wire wire) {
+__host__ __device__ inline Ptr wire_rgt(Wire wire) {
   return wire & 0xFFFFFFFF;
 }
 
@@ -294,29 +294,49 @@ __device__ inline u32 alloc(Worker *worker, Net *net) {
   }
 }
 
-// TODO: use the 4 quad threads
-__device__ inline void split_redex(u64* self_rlen, Wire* self_rbag, u64* targ_rlen, Wire* targ_rbag) {
-  if (*self_rlen > 1 && *self_rlen > *targ_rlen) {
-    u64 init_idx = *targ_rlen;
-    u64 last_idx = *self_rlen;
-    *self_rlen = init_idx;
-    for (u64 i = init_idx; i < last_idx; ++i) {
-      Wire wire = self_rbag[i];
-      self_rbag[i] = mkwire(0,0);
[…]

### 2023-09-11 rust/src/core.rs

@@ -76,7 +76,7 @@ pub struct Book {
 impl Ptr {
   #[inline(always)]
   pub fn new(tag: Tag, val: Val) -> Self {
-    Ptr { data: (((tag as u32) << 28) | (val & 0xFFF_FFF)) }
+    Ptr { data: (((tag as u32) << 28) | (val & 0xFFF_FFFF)) }
   }
 
   #[inline(always)]

### 2023-09-11 rust/src/lang.rs

@@ -391,7 +391,7 @@ pub fn do_alloc_ltree(net: &mut Net, tree: &LTree) -> Ptr {
 pub fn readback_ltree(net: &Net, ptr: Ptr, parent: Parent, vars: &mut HashMap<Parent,String>, fresh: &mut usize) -> LTree {
   match ptr.tag() {
     NIL => {
-      LTree::Era
+      LTree::Var { nam: "?".to_string() }
     },
     ERA => {
       LTree::Era

### 2023-09-11 rust/src/main.rs

@@ -191,7 +191,7 @@ fn main() {
   // ex1 = ((n g_s) g_z)
   define(book, "ex1", "
     $ root
-    & @c23 ~ (0 @g_s (0 @g_z root))
+    & @c20 ~ (0 @g_s (0 @g_z root))
   ");
 
   // This example decreases a binary counter until it reaches 0. It uses recursion, based on
@@ -215,14 +215,14 @@ fn main() {
   "); 
 
   // Initializes the net
-  let net = &mut Net::new(1 << 26);
-  net.boot(book, name_to_u32("ex2"));
+  let net = &mut Net::new(1 << 28);
+  net.boot(book, name_to_u32("ex1"));
 
   // Computes its normal form
   net.normal(book);
 
-  // Shows results and stats
-  println!("[net]\n{}", show_net(&net));
+  //Shows results and stats
+  //println!("[net]\n{}", show_net(&net));
   //println!("{}", net.head.iter().map(|x| format!("{:016x}", x.data)).collect::<Vec<String>>().join(" "));
   println!("size: {}", net.node.len());
   println!("used: {}", net.used);

### 2023-09-11 cuda/hvm2.cu
CUDA local reductions - WIP

@@ -14,20 +14,18 @@ typedef unsigned long long int u64;
 // -------------
 
 // This code is initially optimized for nVidia RTX 4090
-const u32 BLOCK_LOG2    = 8;                                    // log2 of block size
-const u32 BLOCK_SIZE    = 1 << BLOCK_LOG2;                      // threads per block
-const u32 UNIT_SIZE     = 4;                                    // threads per rewrite unit
-const u32 NODE_SIZE     = 1 << 28;                              // max total nodes (2GB addressable)
-const u32 HEAD_SIZE     = BLOCK_SIZE * BLOCK_SIZE * BLOCK_SIZE; // size of unormalized heads
-const u32 GROUP_SIZE    = BLOCK_SIZE * BLOCK_SIZE;              // size os a group of bags
-const u32 REPEAT_RATE   = 32;                                   // local rewrites per global rewrite
-const u32 MAX_TERM_SIZE = 16;                                   // max number of nodes in a term
-
-const u32 MAX_THREADS   = BLOCK_SIZE * BLOCK_SIZE * BLOCK_SIZE;
-const u32 MAX_UNITS     = MAX_THREADS / UNIT_SIZE;
-const u32 MAX_NEW_REDEX = 8;                                    // max new redexes per rewrite
-const u32 RBAG_SIZE     = 64;                                   // max redexes in unit bag
-const u32 BAGS_SIZE     = MAX_UNITS * RBAG_SIZE;                // size of global redex bag
+const u32 BLOCK_LOG2    = 9;                       // log2 of block size
+const u32 BLOCK_SIZE    = 1 << BLOCK_LOG2;         // threads per block
+const u32 UNIT_SIZE     = 4;                       // threads per rewrite unit
+const u32 NODE_SIZE     = 1 << 28;                 // max total nodes (2GB addressable)
+const u32 GROUP_SIZE    = BLOCK_SIZE * BLOCK_SIZE; // size os a group of bags
+const u32 REPEAT_RATE   = 32;                      // local rewrites per global rewrite
+const u32 MAX_TERM_SIZE = 16;                      // max number of nodes in a term
[…]

### 2023-09-11 rust/src/core.rs

@@ -52,7 +52,6 @@ pub struct Node {
 // - root: a single free wire, used as the entrancy point.
 // - acts: a vector of redexes, updated automatically.
 // - node: a vector of nodes, with main ports omitted.
-// - head: a vector of unomralized heads
 // - used: total nodes currently allocated on the graph.
 // - rwts: total graph rewrites performed inside this net.
 // - next: next pointer to allocate memory (internal).
@@ -61,7 +60,6 @@ pub struct Net {
   pub root: Ptr,
   pub acts: Vec<(Ptr, Ptr)>,
   pub node: Vec<Node>,
-  pub isnf: Vec<bool>,
   pub used: usize,
   pub rwts: usize,
       next: usize,
@@ -145,11 +143,6 @@ impl Ptr {
   }
 
   #[inline(always)]
-  pub fn is_target_normal<'a>(&'a self, net: &'a Net) -> bool {
-    return self.is_var() && self.tag() == VRR || net.isnf[self.val() as usize];
-  }
-
-  #[inline(always)]
   pub fn adjust(&self, locs: &[u32]) -> Ptr {
     unsafe {
       return Ptr::new(self.tag(), if self.has_loc() { *locs.get_unchecked(self.val() as usize) } else { self.val() });
@@ -200,7 +193,6 @@ impl Net {
       root: Ptr::new(NIL, 0),
       acts: vec![],
       node: vec![Node::nil(); size],
-      isnf: vec![false; size],
       next: 0,
       used: 0,
       rwts: 0,
@@ -209,8 +201,8 @@ impl Net {
   }
 
   // Creates a net and boots from a REF.
-  pub fn boot(&mut self, book: &Book, root_id: u32) {
-    self.link(book, Ptr::new(VRR, 0), Ptr::new(REF, root_id));
+  pub fn boot(&mut self, root_id: u32) {
+    self.root = Ptr::new(REF, root_id);
   }
 
   // Allocates a consecutive chunk of 'size' nodes. Returns the index.
@@ -272,28 +264,20 @@ impl Net {
     let mut a = a;
     let mut b = b;
     // Dereferences A
-    if a.is_ref() && (b.is_ctr() || b.is_target_normal(self)) {
-      a = self.deref(book, a, &mut 0);
+    if a.is_ref() && b.is_ctr() {
+      a = self.deref(book, a, Ptr::new(NIL,0), &mut 0);
     }
     // Dereferences B
[…]

### 2023-09-11 rust/src/lang.rs

@@ -215,7 +215,7 @@ pub fn num_to_str(num: usize) -> String {
   let mut num = num + 1;
   let mut str = String::new();
   while num > 0 {
-    let c = ((num % 26) as u8 + b'a' - 1) as char;
+    let c = ((num % 26) as u8 + b'a') as char;
     str.push(c);
     num /= 26;
   }

### 2023-09-11 rust/src/main.rs

@@ -191,7 +191,7 @@ fn main() {
   // ex1 = ((n g_s) g_z)
   define(book, "ex1", "
     $ root
-    & @c20 ~ (0 @g_s (0 @g_z root))
+    & @c22 ~ (0 @g_s (0 @g_z root))
   ");
 
   // This example decreases a binary counter until it reaches 0. It uses recursion, based on
@@ -211,15 +211,19 @@ fn main() {
   define(book, "ex3", "
     $ res
     & @brn ~ (0 dep res)
-    & @c4  ~ (0 @S (0 @Z dep))
+    & @c12 ~ (0 @S (0 @Z dep))
   "); 
 
   // Initializes the net
   let net = &mut Net::new(1 << 28);
-  net.boot(book, name_to_u32("ex1"));
+  net.boot(name_to_u32("ex1"));
 
   // Computes its normal form
+  //net.expand(book, Ptr::new(VRR,0));
   net.normal(book);
+  //net.expand(book, Ptr::new(VRR,0));
+  //net.normal(book);
+  //net.expand(book, Ptr::new(VRR,0));
 
   //Shows results and stats
   //println!("[net]\n{}", show_net(&net));
